Improving disaggregation models of malaria incidence by ensembling non-linear models of prevalence

### 2019-11-07 tim-lucas-model-ensembles.tex
Rework results and discussion to be experiment i and ii

In experiment \emph{i} we compared the baseline to two models that used locally trained machine learning models.
First we used only the predictions from the machine learning models trained on local data (ML\textsubscript{l}).
Secondly, we combined the predictions from the machine learning models and the environmental coviarates in one model (Enviro + ML\textsubscript{l}).

@@

\begin{table}[t!]
\caption{Pearson correlations between observed and predicted values. }
\centering
\begin{tabular}{llrrrr}
Country &  Cross-validation & Enviro &  ML\textsubscript{l} &  Enviro + ML\textsubscript{l} \\
\hline 
 Colombia & Random &  0.59 & 0.59 & \textbf{0.61} \\
 Colombia &  Spatial &  0.12 &  \textbf{0.33} &  0.25\\
 Madagascar & Random &   \textbf{0.70} &  0.68 & 0.69 \\
 Madagascar &  Spatial &  0.22 &  \textbf{0.55} & 0.18\\
 Senegal &  Random &  \textbf{0.58} &  0.51 & 0.57 \\
 Senegal &  Spatial &  0.63 &  0.51 &  0.63 \\
 Indonesia &  Random &  0.52 &  0.48 &  \textbf{0.59} \\
 Indonesia &  Spatial &  0.45 &  0.44 &  \textbf{0.51} \\
\end{tabular}
\label{t:results}
\end{table}

@@

Table \ref{t:results} gives the correlation between observed and held out data (under random and spatially cross-validation) for models that use environmental covariates only, local machine learning models only and a combination of the two.
Many of the differences in performance are rather marginal.
Overall, in only one case (random cross-validation in Senegal) is Enviro the best performing model.
In two cases, Enviro perfoms equally well as another model: under spatially cross-validation in Senegal Enviro and Envir + ML\textsubscript{l} perform equally well, under random cross-validation in Colombia Enviro and ML\textsubscript{l} perform equally well.
Of the remaining five cases, in three cases ML\textsubscript{l} performs best and in two cases (random and spatial cross-validation in Indonesia) Envir + ML\textsubscript{l} performs best.
The greatest benefits in predictive performance occur under spatial cross-validation and in the cases when the Enviro model is particularly poor.

@@

Table \ref{t:results2} gives the correlation between observed and held out data (under random and spatially cross-validation) for models that use environmental covariates only, global machine learning models only and a combination of the local and global machine learning models.
In five cases, Enviro is the best or tied best performing model.
In only one case in the ML\textsubscript{g} model the best.
Comparing across Tables \ref{t:results} and \ref{t:results2} we can see that ML\textsubscript{g} is never the best performing model.
In only three cases (Spatial Colombia, Spatial Senegal and Spatial Indonesia) does ML\textsubscript{g} outperform ML\textsubscript{l}.

\begin{table}[t!]
\caption{Pearson correlations between observed and predicted values. }
\centering
\begin{tabular}{llrrrr}
Country &  Cross-validation & Enviro &   ML\textsubscript{g} & ML\textsubscript{l} + ML\textsubscript{g} \\
\hline 
 Colombia & Random &  \textbf{0.59} &0.55 & 0.58 \\
 Colombia &  Spatial &  0.12 &  0.12 & \textbf{0.33}\\
 Madagascar &  Random &  \textbf{0.70} &  0.67 & 0.68 \\
 Madagascar &  Spatial &  0.22 &  0.51 & \textbf{0.55}\\
 Senegal &  Random &  \textbf{0.58} &  0.50 & 0.49 \\
 Senegal &  Spatial &   \textbf{0.63} & 0.55 & 0.52 \\
 Indonesia &  Random &  \textbf{0.52} & 0.32 & 0.46 \\
 Indonesia &  Spatial &  0.45 & 0.41 & 0.45 \\
\end{tabular}
\label{t:results2}
\end{table}
Figures~\ref{f:scatter1} shows scatter plots of the model performance under random cross-validation for all countries and all models.
Figures~\ref{f:scatter2} shows scatter plots of the model performance under spatial cross-validation for all countries and all models.
It can be seen that without environmental covariates, the models in Madagascar fail to predict very high or very low values correctly.
Figure~\ref{f:map} shows the input data and out-of-sample predictions of the ML\textsubscript{l} model in Colombia.
Table~\ref{t:coverage} shows the out-of-sample coverage of the 80\% credible intervals for all models, countries and cross-validation schemes.
The coverage in Colombia is very poor with no models achieving a coverage above 0.4.
Otherwise the coverage is generally acceptable with most values lying between 0.7 and 0.9.
Overall there is a general tendancy for models to be slightly over confident.

We can examine the relationship between the RMSE of the machine learning models to their fitted regression coefficients (weights).
These values are given in Table~\ref{t:mlresults} (with the RMSE of the global models being copied four times).
In all five sets of machine learning models, Random Forest performs the best.
We have not forced $\beta_i > 0$ but we have set the priors for these coefficients with a positive mean.
In nearly all cases the fitted values are positive.
If the prevalence data and incidence data are not biased relative to each other we would expect the models with the lowest RMSE to also have the biggest regression coefficient.
This occurs in three cases where Random Forest has the lowest RMSE and the biggest coefficient.
In a further two cases, Random Forest has the lowest RMSE but GBM has the highest coefficient.
The predictions from Random Forest and GBM are highly correlated in part because they both perform well and in part because they are both tree based models.
Finally, it can be seen that relationship between RMSE and regression coefficients is much weaker in Indonesia.
For the models trained on local data a neural network has the highest fitted coefficient while for the models trained on global data an elastic net has the highest fitted coefficient.
These model both produce very smooth estimates relative to the tree based methods.

@@

\begin{figure}
\centering
\includegraphics[width = 1\textwidth]{figs/col_comparison_map.png} %\caption{Indonesia spatial crossvalidation} 
\caption{
  Left: Observed data for Colombia (grey for zero incidence). Middle: Out-of-sample predictions for the spatial cross-validation, environmental covariates only model. Right: Out-of-sample predictions for the spatial cross-validation, local machine learning only model. For each cross-validation fold, predictions are made for the held out data which are then combined to make a single surface.
}
\label{f:map}
\end{figure}

\begin{table}[t!]
\caption{Covarage of 80\% credible intervals. Values outside 0.7--0.9 are shown in bold.}
\centering
\begin{tabular}{llrrrrr}
Country &  CV & Enviro & ML\textsubscript{l} &  Enviro + ML\textsubscript{l} & ML\textsubscript{g} & ML\textsubscript{l} + ML\textsubscript{g} \\
\hline 
 Colombia & Random & \textbf{0.28} & \textbf{0.28} & \textbf{0.29} & \textbf{0.28} & \textbf{0.30} \\
 Madagascar &  Random & 0.80 & 0.77& 0.75& 0.77& 0.76 \\
 Senegal & Random &0.79 & 0.79& 0.79& 0.79& 0.82 \\
 Indonesia & Random &0.80 & 0.81& 0.78& 0.79& 0.77  \\
 Colombia &  Spatial & \textbf{0.30} & \textbf{0.33}  & \textbf{0.33} & \textbf{0.33} & \textbf{0.34}  \\
 Madagascar & Spatial & \textbf{0.65} & 0.74& 0.70& 0.75 & 0.76   \\
 Senegal & Spatial & 0.85 & \textbf{0.91}& 0.85& \textbf{0.94} & 0.85  \\
  Indonesia & Spatial & 0.80 & 0.78& 0.78& 0.76& 0.75  \\
\end{tabular}
\label{t:coverage}
\end{table}

\begin{table}[t!]
\caption{Machine learning model results and fitted parameters (i.e.\thinspace model weights) of the machine learning predictions only models (local only and global only). }
\centering
\small
\begin{tabular}{ll|rr|rr}
     Country          & Model &      RMSE\textsubscript{l} & $\beta_l$ & RMSE\textsubscript{g} & $\beta_g$ \\ \hline
Madagascar & RF & \textbf{0.100} & 0.538 &  \textbf{0.169} & \textbf{0.529}\\
Madagascar & GBM & 0.105 & \textbf{0.570} & 0.178& 0.432 \\
Madagascar & enet & 0.116 & 0.301 &0.233 & 0.262 \\
Madagascar & nnet & 0.113 & 0.033 &0.220 & 0.364 \\
Madagascar & ppr & 0.109 & 0.469 & 0.205 &  0.403\vspace{0.3cm}\\ 
Colombia & RF & \textbf{0.068} & 0.625 &  \textbf{0.169} & 0.180\\
Colombia & GBM & 0.073 & \textbf{0.952} & 0.178& -0.218  \\
Colombia & enet & 0.070 & 0.219 &0.233 & 0.183 \\
Colombia & nnet & 0.070 & 0.129 &0.220 & 0.527 \\
Colombia & ppr & 0.070 & 0.667 & 0.205 &  \textbf{0.546}\vspace{0.3cm}\\
Senegal & RF & \textbf{0.092} & \textbf{0.339} & \textbf{0.169} & \textbf{0.425} \\
Senegal & GBM & 0.099 & 0.261& 0.178& 0.408 \\
Senegal & enet& 0.103 & 0.344  &0.233 & 0.205 \\
Senegal & nnet & 0.099 & 0.254 &0.220 & 0.190 \\
Senegal & ppr & 0.098 & 0.268& 0.205 &  0.126\vspace{0.3cm}\\
Indonesia & RF& \textbf{0.081} & 0.447 & \textbf{0.169} & 0.178\\
Indonesia & GBM & 0.085 & 0.357 & 0.178& 0.289 \\
Indonesia & enet & 0.091 & 0.303 &0.233 & \textbf{0.526} \\
Indonesia & nnet & 0.089 & \textbf{0.506} &0.220 & 0.316 \\
Indonesia & ppr & 0.089 & 0.364 & 0.205 &  0.089\vspace{0.3cm}\\

@@

Overall, our experiments suggest that using predictions from machine learning models trained on prevalence points provides more accurate predictions than using environmental covariates when fitting disaggregation models of malaria incidence.
This increased performance comes despite the data being on different scales, the data being measurements of different aspects of malaria transmission and despite the imperfect model we have used to translate between the two scales.
However, there is no clear best model between ML\textsubscript{l} and Enviro + ML\textsubscript{l}.
Therefore, when using these methods, both of these models should be fitted and the most predictive model selected for a given dataset.
Furthermore, many of the performance improvements are rather marginal. 
However, in a few cases such as Colombia and Madagascar under spatial cross-validation the performance boost is large.
For example, using just environmental covariates in Colombia under spatial cross-validation gives a correlation between observed and predicted data of 0.12.
Such a model would be unusable for any applied or policy work.
In contrast, using the predictions from the machine learning models trained on local data gives a correlation value of 0.33, which while still poor is possibly useable.

@@

Overall, the models fitted using predictions from machine learning models trained on the global database of prevalence point-surveys were less good than those using predictions from machine learning models trained on local data.
This in itself is not particularly surprising.
However, it does indicate that continental scale models should consider using a mosaic of locally trained machine learning models for example.
What is more surprising is that using both the ML\textsubscript{l} and ML\textsubscript{g} covariates together did not improve performance given that this approach of creating diverse predictions is a core element of stacking methodology.

>>>>

In experiment \emph{i} we compared the baseline to two models that used locally trained machine learning models.
First we used only the predictions from the machine learning models trained on local data (subsequantly called ML\textsubscript{l}).
Secondly, we combined the predictions from the machine learning models and the environmental coviarates in one model (Enviro + ML\textsubscript{l}).

@@

\begin{table}[t!]
\caption{Pearson correlations between observed and predicted values. }
\centering
\begin{tabular}{llrrrr}
Country &  Cross-validation & Enviro &  Enviro + ML\textsubscript{l} &  ML\textsubscript{l} \\
\hline 
 Colombia & Random &  0.59 & \textbf{0.61}& 0.59  \\
 Colombia &  Spatial &  0.12 &  0.25&  \textbf{0.33} \\
 Indonesia &  Random &  0.52 &  \textbf{0.59}&  0.48  \\
 Indonesia &  Spatial &  0.45 &  \textbf{0.51}&  0.44  \\
 Madagascar & Random &   \textbf{0.70}& 0.69 &  0.68  \\
 Madagascar &  Spatial &  0.22 & 0.18&  \textbf{0.55} \\
 Senegal &  Random &  \textbf{0.58} & 0.57&  0.51  \\
 Senegal &  Spatial &  0.63 &  0.63&  0.51  \\
\end{tabular}
\label{t:results}
\end{table}

@@

Table \ref{t:results} gives the correlation between observed and held out data (under random and spatially cross-validation) for experiment \emph{i} (models Enviro, ML\textsubscript{l} and Envir + ML\textsubscript{l}).
Many of the differences in performance are rather marginal.
Enviro is the best performing model in two cases (random cross-validation in Madagascar and Senegal).
In one case, Enviro perfoms equally well as another model; under spatial cross-validation in Senegal Enviro and Enviro + ML\textsubscript{l} perform equally well.
Of the remaining five cases, in two cases ML\textsubscript{l} performs best and in three cases Envir + ML\textsubscript{l} performs best.
The greatest benefits in predictive performance occur under spatial cross-validation and in the cases when the Enviro model is particularly poor.

@@

Figures~\ref{f:scatter1} shows scatter plots of the model performance under random cross-validation for experiment \emph{i}.
Figures~\ref{f:scatter2} shows scatter plots of the model performance under spatial cross-validation for experiment \emph{i}.
It can be seen that without environmental covariates, the models in Madagascar fail to predict very high or very low values correctly.
Figure~\ref{f:map} shows the input data and spatially out-of-sample predictions of the Enviro model and ML\textsubscript{l} model in Colombia.

Table \ref{t:results2} gives the correlation between observed and held out data (under random and spatially cross-validation) for experiment \emph{ii} (models Enviro, ML\textsubscript{g} and ML\textsubscript{l} + ML\textsubscript{g}).
In six cases, Enviro is the best or tied best performing model.
The ML\textsubscript{g} model is never the best performing model and only outpeforms Enviro in one case (spatial cross-validation in Madagascar).
In two cases ML\textsubscript{l} + ML\textsubscript{g} is the best performing model (spatial Colombia and spatial Madagascar).
Comparing across Tables \ref{t:results} and \ref{t:results2} we can see that ML\textsubscript{g} outpeforms ML\textsubscript{l} once (spatial cross-validation in Senegal).
In only two cases (Spatial Senegal and Spatial Indonesia) does ML\textsubscript{l} + ML\textsubscript{g} outperform ML\textsubscript{l}.

\begin{figure}[t!]
\centering
\includegraphics[width = 1\textwidth]{figs/col_comparison_map.png} %\caption{Indonesia spatial crossvalidation} 
\caption{
  A) Observed data for Colombia (grey for zero incidence). B) Out-of-sample predictions for the spatial cross-validation, environmental covariates only model. C) Out-of-sample predictions for the spatial cross-validation, local machine learning only model. For each cross-validation fold, predictions are made for the held out data which are then combined to make a single surface.
}
\label{f:map}
\end{figure}

Table~\ref{t:coverage} shows the out-of-sample coverage of the 80\% credible intervals for all models, countries and cross-validation schemes.
The coverage in Colombia is very poor with no models achieving a coverage above 0.4.
Otherwise the coverage is generally acceptable with most values lying between 0.7 and 0.9.
Overall there is a general tendancy for models to be slightly over confident.

We can examine the relationship between the RMSE of the machine learning models to their fitted regression coefficients (weights).
These values are given in Table~\ref{t:mlresults}.
In all five sets of machine learning models (four sets trained on local data and one set trained on local data), Random Forest performs the best.
We have not forced $\beta_i > 0$ but we have set the priors for these coefficients with a positive mean.
In nearly all cases the fitted values are positive.
If the prevalence data and incidence data are not biased relative to each other we would expect the models with the lowest RMSE to also have the biggest regression coefficient.
This occurs in three cases where Random Forest has the lowest RMSE and the biggest coefficient.
In a further two cases, Random Forest has the lowest RMSE but GBM has the highest coefficient.
The predictions from Random Forest and GBM are highly correlated in part because they both perform well and in part because they are both tree based models.
Finally, it can be seen that the relationship between RMSE and regression coefficients was much weaker in Indonesia.
For the models trained on local data a neural network has the highest fitted coefficient while for the models trained on global data an elastic net has the highest fitted coefficient.

@@

\begin{table}[t!]
\caption{Pearson correlations between observed and predicted values. }
\centering
\begin{tabular}{llrrrr}
Country &  Cross-validation & Enviro &   ML\textsubscript{g} & ML\textsubscript{l} + ML\textsubscript{g} \\
\hline 
 Colombia & Random &  \textbf{0.59} &0.55 & 0.58 \\
 Colombia &  Spatial &  0.12 &  0.12 & \textbf{0.33}\\
 Indonesia &  Random &  \textbf{0.52} & 0.32 & 0.46 \\
 Indonesia &  Spatial &  0.45 & 0.41 & 0.45 \\
 Madagascar &  Random &  \textbf{0.70} &  0.67 & 0.68 \\
 Madagascar &  Spatial &  0.22 &  0.51 & \textbf{0.55}\\
 Senegal &  Random &  \textbf{0.58} &  0.50 & 0.49 \\
 Senegal &  Spatial &   \textbf{0.63} & 0.55 & 0.52 \\

\end{tabular}
\label{t:results2}
\end{table}

\begin{table}[t!]
\caption{Covarage of 80\% credible intervals. Values outside 0.7--0.9 are shown in bold.}
\centering
\begin{tabular}{llrrrrr}
Country &  CV & Enviro & ML\textsubscript{l} &  Enviro + ML\textsubscript{l} & ML\textsubscript{g} & ML\textsubscript{l} + ML\textsubscript{g} \\
\hline 
 Colombia & Random & \textbf{0.28} & \textbf{0.28} & \textbf{0.29} & \textbf{0.28} & \textbf{0.30} \\
 Colombia &  Spatial & \textbf{0.30} & \textbf{0.33}  & \textbf{0.33} & \textbf{0.33} & \textbf{0.34}  \\
 Indonesia & Random &0.80 & 0.81& 0.78& 0.79& 0.77  \\
 Indonesia & Spatial & 0.80 & 0.78& 0.78& 0.76& 0.75  \\
 Madagascar &  Random & 0.80 & 0.77& 0.75& 0.77& 0.76 \\
 Madagascar & Spatial & \textbf{0.65} & 0.74& 0.70& 0.75 & 0.76   \\
 Senegal & Random &0.79 & 0.79& 0.79& 0.79& 0.82 \\
 Senegal & Spatial & 0.85 & \textbf{0.91}& 0.85& \textbf{0.94} & 0.85  \\
\end{tabular}
\label{t:coverage}
\end{table}

\begin{table}[t!]
\caption{Machine learning model results and fitted parameters (i.e.\thinspace model weights) of the machine learning predictions only models (local only and global only). }
\centering
\small
\begin{tabular}{ll|rr|rr}
     Country          & Model &      RMSE\textsubscript{l} & $\beta_l$ & RMSE\textsubscript{g} & $\beta_g$ \\ \hline
Colombia & RF & \textbf{0.068} & 0.625 &  \textbf{0.169} & 0.180\\
Colombia & GBM & 0.073 & \textbf{0.952} & 0.178& -0.218  \\
Colombia & enet & 0.070 & 0.219 &0.233 & 0.183 \\
Colombia & nnet & 0.070 & 0.129 &0.220 & 0.527 \\
Colombia & ppr & 0.070 & 0.667 & 0.205 &  \textbf{0.546}\vspace{0.3cm}\\
Indonesia & RF& \textbf{0.081} & 0.447 & \textbf{0.169} & 0.178\\
Indonesia & GBM & 0.085 & 0.357 & 0.178& 0.289 \\
Indonesia & enet & 0.091 & 0.303 &0.233 & \textbf{0.526} \\
Indonesia & nnet & 0.089 & \textbf{0.506} &0.220 & 0.316 \\
Indonesia & ppr & 0.089 & 0.364 & 0.205 &  0.089\vspace{0.3cm}\\
Madagascar & RF & \textbf{0.100} & 0.538 &  \textbf{0.169} & \textbf{0.529}\\
Madagascar & GBM & 0.105 & \textbf{0.570} & 0.178& 0.432 \\
Madagascar & enet & 0.116 & 0.301 &0.233 & 0.262 \\
Madagascar & nnet & 0.113 & 0.033 &0.220 & 0.364 \\
Madagascar & ppr & 0.109 & 0.469 & 0.205 &  0.403\vspace{0.3cm}\\ 
Senegal & RF & \textbf{0.092} & \textbf{0.339} & \textbf{0.169} & \textbf{0.425} \\
Senegal & GBM & 0.099 & 0.261& 0.178& 0.408 \\
Senegal & enet& 0.103 & 0.344  &0.233 & 0.205 \\
Senegal & nnet & 0.099 & 0.254 &0.220 & 0.190 \\
Senegal & ppr & 0.098 & 0.268& 0.205 &  0.126\\

@@

Overall, experiment \emph{i} suggest that using predictions from machine learning models trained on local prevalence points provides more accurate predictions than using environmental covariates when fitting disaggregation models of malaria incidence.
This increased performance comes despite the data being on different scales, the data being measurements of different aspects of malaria transmission and despite the imperfect model we have used to translate between the two scales.
However, there is no clear best model between ML\textsubscript{l} and Enviro + ML\textsubscript{l}.
Therefore, when using these methods, both of these models should be fitted and the most predictive model selected for a given dataset.
Furthermore, many of the performance improvements are rather marginal. 
However, in a few cases such as Colombia and Madagascar under spatial cross-validation the performance boost is large.
For example, using just environmental covariates in Madagascar under spatial cross-validation gives a correlation between observed and predicted data of 0.22.
Such a model would be unusable for any applied or policy work.
In contrast, using the predictions from the machine learning models trained on local data gives a correlation value of 0.55, which while still relatively poor is possibly a useful model.

@@

Overall, the models fitted using predictions from machine learning models trained on the global database of prevalence point-surveys were less good than those using either environmental data alone or than those using predictions from machine learning models trained on local data.
This in itself is not particularly surprising.
However, it does indicate that continental scale models should consider using a mosaic of locally trained machine learning models for example.
What is more surprising is that using both the ML\textsubscript{l} and ML\textsubscript{g} covariates together did not improve performance given that this approach of creating diverse predictions is a core element of stacking methodology.
